---
title: "DSC 011 S26 — Lecture 29 Demo + Lab 10"
subtitle: "Reproducible Project Setup: Tools, Scripts, Data, and Distribution"
author: "FirstName LastName"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: false
    theme: cosmo
    embed-resources: true
    include-in-header:
      text: |
        <style>
        kbd {
          display: inline-block;
          padding: 2px 6px;
          font-size: 0.85em;
          font-family: monospace;
          background-color: #f5f5f5;
          border: 1px solid #ccc;
          border-bottom: 2px solid #aaa;
          border-radius: 3px;
        }
        </style>
execute:
  echo: true
  warning: false
  message: false
  error: true
urlcolor: blue
linkcolor: blue
engine: knitr
---

```{r setup, include=FALSE, cache=FALSE, message=FALSE, warning=FALSE, results='hide'}
options(digits = 5)
knitr::opts_chunk$set(
  echo     = TRUE,
  cache    = FALSE,
  warning  = FALSE,
  message  = FALSE,
  collapse = FALSE,
  comment  = "",
  error    = TRUE    # never halt rendering on bash exit code 1
)

# ── Unset stale virtual environment from other projects ───────────────────────
Sys.unsetenv("VIRTUAL_ENV")

# ── Fix PATH for bash chunks on macOS (safe no-op on Linux/WSL2) ─────────────
# RStudio launched from the Dock inherits a minimal launchd PATH that omits
# /opt/homebrew/bin. This prepends Homebrew paths so knitr bash chunks see
# bash 5 and system tools like check_bash.
homebrew_paths <- c("/opt/homebrew/bin", "/usr/local/bin")
existing <- homebrew_paths[dir.exists(homebrew_paths)]
if (length(existing) > 0) {
  old_path <- Sys.getenv("PATH")
  new_path <- paste(c(existing, old_path), collapse = ":")
  Sys.setenv(PATH = new_path)
}

# ── R check infrastructure ────────────────────────────────────────────────────
# source("dsc011_check_v2.2.R")   # uncomment after wget delivers this file
```

---

## Preliminaries

### Instructions for This Notebook

This is a **single notebook covering two submissions:**

- **Lecture 29 Demo** — complete Steps 1–7 during class. Submit your
  rendered HTML by **2:30 pm today (April 9)**. Late submissions accepted
  until **11:59 pm April 10**.
- **Lab 10** — complete Steps 8–16 independently in your lab section.
  Submit by **5:30 pm Wednesday / 4:30 pm Friday**. Late submissions
  accepted until **Friday April 11 at 11:59 pm**.

You render and submit **the same `.qmd` file twice** — once at the Lecture
29 submission point (after Step 7), and again when Lab 10 is complete.

::: {.callout-important}
### Complete Steps 1–5 in your terminal before opening this notebook

This notebook must render from inside the project directory. Open
`DSC-011-Sp26-Baby-Names.Rproj` only after you have created the project
directory in Step 5. The render guard below will tell you if anything is
wrong — it never halts rendering, it just prints a clear diagnostic.
:::

::: {.callout-tip}
### Command-line habits for this class

- **Tab-complete** directory and file names in the terminal. If nothing
  completes, you have a typo or are in the wrong directory.
- **Copy-paste** paths from this notebook rather than typing them.
  WSL2 Windows Terminal: <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>C</kbd> /
  <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>V</kbd>.
  macOS: <kbd>Cmd</kbd>+<kbd>C</kbd> / <kbd>Cmd</kbd>+<kbd>V</kbd>.
- **Develop in the terminal, memorialize in the notebook.** Write and test
  commands in your shell first. Once they work, paste them into the chunk.
- **Read errors before asking for help** — they almost always name the
  problem exactly.
- **In UNIX, silence is golden.** No output + exit code 0 = success.
:::

::: {.callout-note}
### Role Responsibilities

- **All Team Members** work individually to follow and reproduce the demo.
- **All Team Members** signal the Reporter or Scrum Master when blocked.
- **Scrum Master** leads the stand-up (~2 min), timeboxes work, signals
  instructor when team is ready or needs help.
- **Reporter** records blockages and resolutions in the log tables below,
  and summarizes the retrospective at the end.
- **Developers** test and troubleshoot, help blocked teammates.
:::

---

## Render Guard

This chunk runs on every render. It never halts rendering — instead it
prints `✓` or `✗ INCORRECT` with a clear fix message for each check.
Read it carefully before proceeding with any step.

```{bash render-guard, error=TRUE}
echo "=== Render Guard ==="
echo ""

# ── 1. Working directory ──────────────────────────────────────────────────────
EXPECTED="$HOME/DSC-011-Sp26/DSC-011-Sp26-Baby-Names"
ACTUAL=$(pwd)
if [[ "$ACTUAL" == "$EXPECTED" ]]; then
    echo "✓ Working directory: $ACTUAL"
else
    echo "✗ INCORRECT — Wrong working directory"
    echo "  Rendering from: $ACTUAL"
    echo "  Expected:       $EXPECTED"
    echo "  Fix:"
    echo "    1. Close this project (File → Close Project)"
    echo "    2. cd ~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names"
    echo "    3. Open DSC-011-Sp26-Baby-Names.Rproj"
    echo "    4. Re-render"
fi

echo ""

# ── 2. bash version ───────────────────────────────────────────────────────────
BASH_MAJOR="${BASH_VERSINFO[0]}"
BASH_MINOR="${BASH_VERSINFO[1]}"
if [[ "$BASH_MAJOR" -gt 5 ]] || \
   { [[ "$BASH_MAJOR" -eq 5 ]] && [[ "$BASH_MINOR" -ge 1 ]]; }; then
    echo "✓ bash $BASH_VERSION at $(which bash)"
else
    echo "✗ INCORRECT — bash $BASH_VERSION is too old (need 5.1+)"
    echo "  macOS fix:"
    echo "    brew install bash"
    echo "    echo /opt/homebrew/bin | sudo tee /etc/paths.d/homebrew"
    echo "    Quit RStudio fully; relaunch: open -a RStudio"
    echo "  WSL2 fix:"
    echo "    sudo apt update && sudo apt install --only-upgrade bash"
fi

echo ""

# ── 3. check_bash ─────────────────────────────────────────────────────────────
if command -v check_bash &>/dev/null; then
    echo "✓ $(check_bash --version) at $(which check_bash)"
else
    echo "✗ INCORRECT — check_bash not found on PATH"
    echo "  Fix: complete Step 3 (clone repo, just install-system)"
    echo "  then re-render."
fi

echo ""

# ── 4. git available ──────────────────────────────────────────────────────────
if command -v git &>/dev/null; then
    echo "✓ $(git --version)"
else
    echo "✗ INCORRECT — git not found"
    echo "  macOS: brew install git"
    echo "  WSL2:  sudo apt install git"
fi

echo ""
echo "=== End Render Guard ==="
```

---

# LECTURE 29 DEMO

## Background: Why We Are Rebuilding from Scratch

Demo 28 revealed three infrastructure problems that caused nearly all
student errors. Today we fix all three permanently before continuing to
build the Baby Names project pipeline.

**Problem 1 — `check_bash` in `~/bin/`** (caused 63% of INCORRECT/Error
outcomes). knitr bash chunks run as non-interactive subprocesses that never
source `~/.bashrc`, so `~/bin/` is invisible to them. Fix: install
`check_bash` system-wide at `/usr/local/bin/` using `just install-system`.

**Problem 2 — Wrong working directory** (56% of students). The notebook
landed in `Downloads/`, `Desktop/`, or paths with typos because students
placed it manually. Fix: deliver all notebooks via `wget` directly into
the project directory, and use a non-halting render guard that diagnoses
problems clearly without stopping the render.

**Problem 3 — Orphaned `~/DSC011/` path.** The new canonical location for
all course project work is `~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names`. The
old path is retired — leave it in place (it has your git history), but
do not use it for new work.

---

## Step 1: Install `just` and `glow`

**`just`** is a modern command runner — like `make` but purpose-built for
running named project commands defined in a `Justfile`. We use it today to
install `check_bash` in a readable, auditable, reproducible way.

**`glow`** renders Markdown beautifully in your terminal. Instead of
opening GitHub in a browser to read a README before installing something,
you clone the repo and read its docs at the command line — staying in the
terminal workflow throughout.

Run the appropriate commands for your platform in your **terminal**:

```
# macOS
brew install just glow

# WSL2/Ubuntu — just
sudo apt update && sudo apt install just

# WSL2/Ubuntu — glow (via Charm apt repo)
sudo mkdir -p /etc/apt/keyrings
curl -fsSL https://repo.charm.sh/apt/gpg.key \
    | sudo gpg --dearmor -o /etc/apt/keyrings/charm.gpg
echo "deb [signed-by=/etc/apt/keyrings/charm.gpg] \
    https://repo.charm.sh/apt/ * *" \
    | sudo tee /etc/apt/sources.list.d/charm.list
sudo apt update && sudo apt install glow
```

Verify:

```{bash verify-just-glow, error=TRUE}
echo "just:  $(just --version 2>/dev/null || echo 'NOT FOUND')"
echo "glow:  $(glow --version 2>/dev/null | head -1 || echo 'NOT FOUND')"
```

#### Reporter's Block/Resolution Log — Step 1

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 2: Install `firefox`, `w3m`, and `wslu`

**`firefox`** lets you open rendered HTML output from the terminal with a
single command after `quarto render`.

**`w3m`** is a terminal-based web browser — it renders HTML directly in
your terminal without opening any GUI window. It is useful for quickly
inspecting rendered output without leaving your shell, and works on any
platform including WSL2 configurations where GUI apps are not yet
available.

**`wslu`** (WSL2 only) provides `wslview`, which opens any file or URL in
your Windows default browser directly from the WSL2 terminal, handling the
Linux→Windows path translation automatically.

**macOS:**

```
# Skip firefox if already installed at /Applications/Firefox.app
brew install --cask firefox
brew install w3m
```

**WSL2/Ubuntu:**

```
sudo apt update && sudo apt install firefox w3m wslu
```

::: {.callout-note}
### Firefox on WSL2 requires WSLg

WSL2 Firefox opens a native Linux GUI window using WSLg (Windows
Subsystem for Linux GUI), which you demonstrated in an earlier session.
Verify WSLg is active: `echo $DISPLAY` should return something like `:0`.
If empty, run `wsl --shutdown` in PowerShell and reopen WSL2.

`wslview` opens the file in your existing Windows browser without
requiring WSLg. `w3m` is the most robust of all — it requires no GUI
and works everywhere.
:::

### Viewing rendered HTML: four methods

| Method | Command | Works on |
|---|---|---|
| `w3m` | `w3m report.html` | macOS + WSL2, no GUI needed |
| `wslview` | `wslview report.html` | WSL2, opens Windows browser |
| Linux Firefox | `firefox report.html &` | WSL2 with WSLg active |
| macOS open | `open -a Firefox report.html` | macOS only |

```{bash verify-firefox-w3m-wslu, error=TRUE}
echo "firefox: $(firefox --version 2>/dev/null | head -1 || echo 'NOT FOUND')"
echo "w3m:     $(w3m -version 2>/dev/null | head -1 || echo 'NOT FOUND')"
# WSL2 students: uncomment the next line
# echo "wslview: $(wslview --version 2>/dev/null | head -1 || echo 'NOT FOUND')"
```

#### Reporter's Block/Resolution Log — Step 2

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 3: Clone `dsc011-check-bash`, Read the README, and Install

This is the canonical workflow for installing any well-documented
command-line tool: **clone → read → verify → install**. We practice the
full sequence so you understand every step.

### 3a. Clone the repository

```
cd ~
git clone https://github.com/dhard/dsc011-check-bash.git
cd dsc011-check-bash
ls
```

You should see: `check_bash`, `install_check_bash.sh`, `test_check_bash.sh`,
`Justfile`, `README.md`, `CHANGELOG.md`, `CONTRIBUTING.md`.

### 3b. List the available `just` recipes

```
just --list
```

You will see: `install`, `install-system`, `uninstall`, `test`, `check`,
`version`. Each recipe is a named command defined in the `Justfile`. Run
`cat Justfile` to see exactly what each recipe does.

### 3c. Read the README with `glow`

```
glow README.md
```

Use arrow keys to scroll, <kbd>q</kbd> to quit. Find the **Typical
workflow** section — it shows exactly the sequence you are about to run.

### 3d. Uninstall any previous `check_bash`

`just uninstall` removes `check_bash` from both `~/bin/` and
`/usr/local/bin/` (whichever exist) and cleans the PATH entry from your
shell rc file. Safe to run even if `check_bash` was never installed.

```
just uninstall
```

### 3e. Run the test suite

Before installing, verify that `check_bash` works correctly on your system:

```
just test
```

This runs 89 tests across 15 sections. All 89 should pass. If any fail,
stop and ask for help before proceeding.

### 3f. Install system-wide

```
just install-system
```

This installs `check_bash` to `/usr/local/bin/` and prompts for your
password if `sudo` is needed.

::: {.callout-note}
### Why `/usr/local/bin/` works where `~/bin/` did not

knitr runs each bash chunk as a fresh **non-interactive** subprocess that
does not source `~/.bashrc` or `~/.zshrc`. It inherits only the PATH that
R has at render time — which comes from however RStudio was launched, not
from your shell startup files. `/usr/local/bin/` is in the **system
default PATH** present in every process on both platforms. `~/bin/` is
added only by your shell startup file, which runs only in interactive
sessions. This is why `check_bash` worked in your terminal but caused
`bash: check_bash: command not found` in 63% of Demo 28 submissions.
:::

```{bash verify-check-bash, error=TRUE}
echo "Location: $(which check_bash 2>/dev/null || echo 'NOT FOUND')"
echo "Version:  $(check_bash --version 2>/dev/null || echo 'NOT FOUND')"
echo ""
if [[ -f "$HOME/bin/check_bash" ]]; then
    echo "⚠ WARNING: check_bash still exists at ~/bin"
    echo "  Run 'just uninstall' from ~/dsc011-check-bash, then reinstall"
else
    echo "✓ check_bash is not in ~/bin (correct — system install only)"
fi
```

#### Reporter's Block/Resolution Log — Step 3

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 4: Clone the Demo 28 Analysis Repo, Render, and View

The Demo 28 submission analysis — a fully reproducible report showing what
went wrong and why — lives in a GitHub repository. Today we clone it,
render it with Quarto, and open the result from the terminal. This is the
`git clone` → `quarto render` → review workflow you will use for your own
analysis deliverables.

### 4a. Clone the analysis repo

```
cd ~
git clone https://github.com/dhard/DSC011-Sp26-Demo28-Analysis.git
cd DSC011-Sp26-Demo28-Analysis
ls
```

::: {.callout-note}
The URL above is a placeholder — the actual URL will be announced in
class and posted to CatCourses.
:::

### 4b. Render the report

```
quarto render demo28_report.qmd
```

### 4c. Open in your browser from the terminal

**All platforms — `w3m` (terminal, no GUI needed):**

```
w3m ~/DSC011-Sp26-Demo28-Analysis/demo28_report.html
```

Press <kbd>q</kbd> to quit `w3m`.

**macOS — Firefox:**

```
open -a Firefox ~/DSC011-Sp26-Demo28-Analysis/demo28_report.html
```

**WSL2 — `wslview` (opens in Windows default browser, no WSLg needed):**

```
wslview ~/DSC011-Sp26-Demo28-Analysis/demo28_report.html
```

**WSL2 — Linux Firefox (requires WSLg):**

```
firefox ~/DSC011-Sp26-Demo28-Analysis/demo28_report.html &
```

```{bash verify-analysis-render, error=TRUE}
f="$HOME/DSC011-Sp26-Demo28-Analysis/demo28_report.html"
if [[ -f "$f" ]]; then
    echo "✓ demo28_report.html rendered ($(du -h "$f" | cut -f1))"
else
    echo "✗ INCORRECT — demo28_report.html not found"
    echo "  cd ~/DSC011-Sp26-Demo28-Analysis && quarto render demo28_report.qmd"
fi
```

#### Reporter's Block/Resolution Log — Step 4

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 5: Create the New Project Directory and Attach an RStudio Project

### 5a. Create the parent and project directories

The `-p` flag creates intermediate directories as needed and does not
error if they already exist:

```
mkdir -p ~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names
cd ~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names
pwd
```

The `pwd` output should end with `.../DSC-011-Sp26/DSC-011-Sp26-Baby-Names`.

::: {.callout-important}
### Old path vs. new path

| | Path |
|---|---|
| **Old (orphaned — do not use)** | `~/DSC011/DSC-011-Sp26-Baby-Names` |
| **New (correct)** | `~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names` |

These differ by one character in the parent. Do not type either path by
hand — copy-paste from this notebook or tab-complete.
:::

### 5b. Attach an RStudio project

An RStudio project (`.Rproj` file) anchors the working directory to the
project root. When you open the `.Rproj` file, RStudio sets the working
directory automatically — this is what makes the render guard work.

1. In RStudio: **File → New Project → Existing Directory**
2. Browse to `~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names`
3. Click **Create Project**

RStudio creates `DSC-011-Sp26-Baby-Names.Rproj` and reopens the project.

```{bash verify-project-dir, error=TRUE}
EXPECTED="$HOME/DSC-011-Sp26/DSC-011-Sp26-Baby-Names"
if [[ -d "$EXPECTED" ]]; then
    echo "✓ Project directory exists: $EXPECTED"
else
    echo "✗ INCORRECT — directory not found"
    echo "  Run: mkdir -p ~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names"
fi
if [[ -f "$EXPECTED/DSC-011-Sp26-Baby-Names.Rproj" ]]; then
    echo "✓ .Rproj file exists"
else
    echo "✗ INCORRECT — .Rproj not found"
    echo "  Create via File → New Project → Existing Directory"
fi
```

#### Reporter's Block/Resolution Log — Step 5

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 6: Fetch This Notebook with `wget` and Render It

All demo and lab notebooks from this point on are delivered via `wget`
directly into the project directory. This guarantees the file lands on
the Linux filesystem in exactly the right location, with correct line
endings, so it renders without error.

### 6a. Fetch the notebook

Run in your **terminal** from inside the project directory:

```
cd ~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names
wget https://COURSE-URL-TBD/DSC011-S26-Lecture29-Demo-Lab10_FirstName_LastName.qmd
```

::: {.callout-note}
The URL is a placeholder — the actual URL will be announced in class
and posted to CatCourses. Always `cd` to the project directory **before**
running `wget` so the file lands in the right place.
:::

### 6b. Personalize, render, and open in your browser

1. Open `DSC-011-Sp26-Baby-Names.Rproj` in RStudio
2. Open this `.qmd` file in the source pane
3. Replace `FirstName LastName` in the YAML `author:` field with your name
4. **Save As** — replace `FirstName_LastName` in the filename with your name
5. Render to HTML
6. Open the rendered HTML from your terminal:

**All platforms — `w3m`:**

```
w3m DSC011-S26-Lecture29-Demo-Lab10_YourName.html
```

**macOS — Firefox:**

```
open -a Firefox DSC011-S26-Lecture29-Demo-Lab10_YourName.html
```

**WSL2 — `wslview`:**

```
wslview DSC011-S26-Lecture29-Demo-Lab10_YourName.html
```

**WSL2 — Linux Firefox:**

```
firefox DSC011-S26-Lecture29-Demo-Lab10_YourName.html &
```

Read the **Render Guard** output at the top of the rendered HTML carefully.
All four checks should show `✓` before you proceed.

```{bash verify-notebook-fetched, error=TRUE}
nb=$(ls DSC011-S26-Lecture29-Demo-Lab10_*.qmd 2>/dev/null | head -1)
if [[ -n "$nb" ]]; then
    echo "✓ Notebook found: $nb"
else
    echo "✗ INCORRECT — notebook .qmd not found in $(pwd)"
    echo "  Run wget from inside ~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names"
fi
```

#### Reporter's Block/Resolution Log — Step 6

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 7: Create the `scripts/` Directory, Empty Script Files, and `data/` Tree

We create all project directories and the three pipeline script files now
as **empty placeholders**. We fill them in during Lab 10. Creating them
first establishes the project structure and lets git track it from the
start.

### 7a. Create `scripts/` and empty script files

```
mkdir -p scripts
touch scripts/00-setup.sh scripts/01-obtain.sh scripts/02-scrub.sh
```

`touch` creates a file if it does not exist, or updates its timestamp if
it does. It is safe to run repeatedly.

### 7b. Create the `data/` directory tree

| Directory | Contents | git-tracked? |
|---|---|---|
| `data/raw/` | Original HTML files from SSA — never modified | No (gitignored) |
| `data/processed/` | Intermediate files (plain text, etc.) | No (gitignored) |
| `data/clean/` | Final cleaned outputs | No (gitignored) |

Data directories are gitignored because their contents are large and
regenerable from the scripts. The **scripts that produce them** are what
we version-control.

```
mkdir -p data/raw data/processed data/clean
```

### 7c. Validate the full scaffold

```{bash verify-scaffold, error=TRUE}
echo "=== Project scaffold ==="
echo ""
tree --charset ascii scripts/ data/ 2>/dev/null
echo ""

# Validate with check_bash
answer=$(tree --charset ascii scripts/ data/ 2>/dev/null)
check_bash "$answer" "HASH_TREE_SCAFFOLD_TBD"
```

::: {.callout-note}
The hash above is a placeholder to be filled in before distribution.
:::

#### Reporter's Block/Resolution Log — Step 7

| Block | Resolution |
|:------|:-----------|
|       |            |

---

::: {.callout-important}
## ★ LECTURE 29 DEMO SUBMISSION POINT

**Render this notebook now and submit the HTML to CatCourses.**

- **Due:** 2:30 pm today, April 9, 2026
- **Late deadline:** 11:59 pm April 10, 2026
- **What to submit:** the rendered `.html` file (not the `.qmd`)

Before submitting, confirm:

- [ ] Your name is in the `author:` field (not `FirstName LastName`)
- [ ] The Render Guard shows `✓` for all four checks
- [ ] Step 3f shows `check_bash` at `/usr/local/bin/`
- [ ] Step 7c shows the directory scaffold with `tree`

A notebook with some `✗ INCORRECT` outputs is still submittable for
partial credit — do not wait for perfection. Submit what you have, then
fix and resubmit before the late deadline.
:::

---

# LAB 10

::: {.callout-important}
### Lab 10 is independent work

Complete Steps 8–16 during your lab section or independently. Submit
**the same `.qmd` file** rendered to HTML again when finished.

- **Due:** 5:30 pm Wednesday April 9 / 4:30 pm Friday April 11
- **Late deadline:** Friday April 11 at 11:59 pm
:::

---

## Step 8: Catch-Up Validation and git Repository Check

Before building further, confirm that your Demo 29 work is complete. The
checks below are non-blocking — `✗ INCORRECT` outputs tell you what to
fix before proceeding.

```{bash lab10-catchup, error=TRUE}
echo "=== Lab 10 catch-up validation ==="
echo ""

# Working directory
EXPECTED="$HOME/DSC-011-Sp26/DSC-011-Sp26-Baby-Names"
[[ "$(pwd)" == "$EXPECTED" ]] \
    && echo "✓ Working directory correct" \
    || echo "✗ Wrong directory: $(pwd)"

# Scripts
for f in scripts/00-setup.sh scripts/01-obtain.sh scripts/02-scrub.sh; do
    [[ -f "$f" ]] && echo "✓ $f" || echo "✗ $f missing"
done

# Data directories
for d in data/raw data/processed data/clean; do
    [[ -d "$d" ]] && echo "✓ $d/" || echo "✗ $d/ missing"
done

# check_bash
command -v check_bash &>/dev/null \
    && echo "✓ check_bash: $(which check_bash)" \
    || echo "✗ check_bash not on PATH — complete Step 3"

# git available
command -v git &>/dev/null \
    && echo "✓ $(git --version)" \
    || echo "✗ git not found"

# git repo initialized (expected: not yet — we do this in Step 9)
if git rev-parse --git-dir &>/dev/null 2>&1; then
    echo "✓ git repo initialized"
    echo "  Commits so far: $(git log --oneline 2>/dev/null | wc -l | tr -d ' ')"
else
    echo "○ git repo not yet initialized (expected — complete Step 9)"
fi
```

#### Reporter's Block/Resolution Log — Step 8

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 9: Initialize git, Create `.gitignore`, and Make the First Commit

### 9a. What is git and why do we use it?

**git** is the world's most widely used version control system. It records
a complete history of every change to your project files — who made each
change and why. For a data science project this means:

- You can always recover any previous version of a script
- You can see exactly what changed between runs and when
- Collaborators can work on the same project without overwriting each other
- The repository is self-documenting: the git log is a lab notebook

Every **commit** is a snapshot of the project at a moment in time. Every
snapshot has a message. The sequence of commits tells the story of how the
project was built.

### 9b. Initialize the repository

```
cd ~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names
git init
git config user.email "YOUREMAIL@ucmerced.edu"
git config user.name "Your Name"
```

### 9c. Create `.gitignore`

The `.gitignore` file tells git which files and directories to leave
untracked. We gitignore data directories (large, regenerable), virtual
environments (machine-specific), and editor backup files.

```
mg .gitignore
```

Paste the following, save
(<kbd>Ctrl</kbd>+<kbd>x</kbd><kbd>Ctrl</kbd>+<kbd>s</kbd>), and quit
(<kbd>Ctrl</kbd>+<kbd>x</kbd><kbd>Ctrl</kbd>+<kbd>c</kbd>):

```
# Data — regenerated by scripts, not committed
data/raw/
data/processed/
data/clean/

# Python virtual environment
.venv/
__pycache__/
*.pyc
.python-version

# R
.Rproj.user/
*.Rhistory
*.RData
renv/library/

# OS and editor
.DS_Store
Thumbs.db
*~
*.orig
```

### 9d. Stage and commit

```
git add .
git status
git commit -m "Initial commit: project scaffold, scripts/, data/, .gitignore"
```

Read `git status` before committing — confirm that `data/raw/`,
`data/processed/`, `data/clean/`, and `.venv/` (if it exists) are **not**
listed (they are gitignored).

```{bash verify-first-commit, error=TRUE}
answer=$(git log --oneline 2>/dev/null | wc -l | tr -d ' ')
echo "Commit count: $answer"
check_bash "$answer" "6b86b273ff34fce19d6b804eff5a3f5747ada4eaa22f1d49c01e52ddb7875b4b"
```

#### Reporter's Block/Resolution Log — Step 9

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 10: Install `uv` and Set Up the Python Environment

### Background: Why `uv`?

Python projects need two things to be reproducible: a specific Python
version and specific versions of every library. Traditional tools (`pip`,
`venv`, `pyenv`, `conda`) each solve part of this problem. `uv` solves
all of it in one fast, modern tool.

In this project, `uv` manages the Python environment that runs `html2text`,
the library we use to convert SSA's HTML files to plain text. Any
collaborator who clones this repo and runs `uv sync` gets an identical
environment — same Python version, same library versions.

::: {.callout-note}
### Check for `pyenv` first

`uv` and `pyenv` conflict. Before installing `uv`, verify:

```
which pyenv       # should return nothing
echo $PYENV_ROOT  # should return nothing
```

If either returns a value, stop and ask the instructor for help.
:::

### 10a. Install `uv`

```
curl -LsSf https://astral.sh/uv/install.sh | sh
```

After installing, source your shell configuration so `uv` is on PATH:

```
source $HOME/.local/bin/env
```

Or open a new terminal tab.

### 10b. Enable shell autocompletion (optional but recommended)

```
# zsh (macOS default shell)
echo 'eval "$(uv generate-shell-completion zsh)"' >> ~/.zshrc
echo 'eval "$(uvx --generate-shell-completion zsh)"' >> ~/.zshrc

# bash (WSL2 default shell)
echo 'eval "$(uv generate-shell-completion bash)"' >> ~/.bashrc
echo 'eval "$(uvx --generate-shell-completion bash)"' >> ~/.bashrc
```

Restart your shell or source the rc file afterward.

### 10c. Verify `uv`

```{bash verify-uv, error=TRUE}
if command -v uv &>/dev/null; then
    echo "✓ $(uv --version)"
    echo "✓ Python via uv: $(uv run python --version 2>/dev/null)"
else
    echo "✗ INCORRECT — uv not found on PATH"
    echo "  Run: curl -LsSf https://astral.sh/uv/install.sh | sh"
    echo "  Then: source \$HOME/.local/bin/env"
fi
```

### 10d. Initialize `uv` in the project and add `html2text`

```
cd ~/DSC-011-Sp26/DSC-011-Sp26-Baby-Names
uv init
uv add html2text
```

`uv init` creates `pyproject.toml` (project metadata and dependencies) and
`uv.lock` (exact pinned versions of every dependency and their
dependencies). `uv add html2text` adds the library as a dependency,
updates both files, and installs it into `.venv/`.

::: {.callout-note}
### What these files do

| File | Purpose | git-tracked? |
|---|---|---|
| `pyproject.toml` | Declares project name, Python version, dependencies | Yes |
| `uv.lock` | Pins exact versions of all dependencies | Yes |
| `.venv/` | The actual virtual environment | No (gitignored) |

`pyproject.toml` and `uv.lock` are committed. `.venv/` is not — it is
regenerated by `uv sync` on any machine. This is how `uv` achieves
reproducibility: the lockfile is the source of truth.
:::

```{bash verify-uv-init, error=TRUE}
[[ -f pyproject.toml ]] \
    && echo "✓ pyproject.toml exists" \
    || echo "✗ pyproject.toml missing — run: uv init"

[[ -f uv.lock ]] \
    && echo "✓ uv.lock exists" \
    || echo "✗ uv.lock missing — run: uv init"

uv run html2text --version &>/dev/null \
    && echo "✓ html2text available via uv run" \
    || echo "✗ html2text not found — run: uv add html2text"
```

#### Reporter's Block/Resolution Log — Step 10

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 11: Download One HTML File by Hand and Understand the Conversion

Before automating anything, we do one step manually: downloading a single
HTML file and converting it to plain text. This teaches you exactly what
the scripts will automate, and what to look for when things go wrong.

### 11a. Download one HTML file

```
wget https://www.ssa.gov/oact/babynames/decades/names1990s.html \
     -O data/raw/names1990s.html
```

The `-O` flag specifies the output filename and path. Always use it with
`wget` so files land exactly where you intend.

Inspect the first 40 lines of raw HTML:

```
head -40 data/raw/names1990s.html
```

Raw HTML is not human-readable as tabular data — the data is buried in
tags. `html2text` strips the tags and produces readable plain text that
standard UNIX tools can parse.

### 11b. Understand the Python one-liner

The conversion command is:

```
uv run python -c \
    'import sys, html2text; print(html2text.html2text(sys.stdin.read()))' \
    < data/raw/names1990s.html \
    > data/processed/names1990s.txt
```

Each part explained:

| Part | What it does |
|---|---|
| `uv run python` | Runs Python from the project `.venv/` — no activation needed |
| `-c '...'` | Passes a Python program as a string on the command line |
| `import sys, html2text` | Loads `sys` (for stdin) and the `html2text` library |
| `sys.stdin.read()` | Reads all of stdin as one string |
| `html2text.html2text(...)` | Converts the HTML string to plain text |
| `print(...)` | Writes the result to stdout |
| `< data/raw/names1990s.html` | Redirects the HTML file into stdin |
| `> data/processed/names1990s.txt` | Redirects stdout to the output file |

### 11c. Run the conversion by hand

```
uv run python -c \
    'import sys, html2text; print(html2text.html2text(sys.stdin.read()))' \
    < data/raw/names1990s.html \
    > data/processed/names1990s.txt
```

Inspect the output:

```
head -30 data/processed/names1990s.txt
```

You should see a table with columns for rank, boys name, boys count,
girls name, and girls count — separated by pipe characters (`|`). This
is the structure the scrub pipeline will parse next week.

```{bash verify-manual-conversion, error=TRUE}
if [[ -f data/processed/names1990s.txt ]]; then
    wc=$(wc -l < data/processed/names1990s.txt | tr -d ' ')
    echo "✓ names1990s.txt exists ($wc lines)"
    echo ""
    echo "First 5 lines:"
    head -5 data/processed/names1990s.txt
else
    echo "✗ INCORRECT — names1990s.txt not found"
    echo "  Run the uv run python conversion above"
fi
```

Count how many data lines this file contains (lines starting with a rank
number):

```{bash count-data-lines-single, error=TRUE}
answer=$(grep -c "^ *[0-9]" data/processed/names1990s.txt 2>/dev/null \
         | tr -d ' ')
echo "Data lines in names1990s.txt: $answer"
echo "(Each decade has 200 ranks — 100 boys + 100 girls)"
```

#### Reporter's Block/Resolution Log — Step 11

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 12: Write and Run `00-setup.sh`

`00-setup.sh` is the **environment bootstrap script**. Any collaborator
who clones this repository runs it once to recreate the Python environment
and the data directory scaffold. It is the entry point for reproducibility.

### 12a. Open and write the script

```
mg scripts/00-setup.sh
```

```bash
#!/usr/bin/env bash
set -euo pipefail

# 00-setup.sh — Bootstrap the project environment
# DSC 011 Sp26 — DSC-011-Sp26-Baby-Names
#
# Run once after cloning the project or setting up a new machine.
# Creates the data directory scaffold and installs the Python environment.
#
# Usage (from any directory):
#   bash scripts/00-setup.sh

PROJ_ROOT="$(cd "$(dirname "${BASH_SOURCE[0]}")/.." && pwd)"
cd "$PROJ_ROOT"

echo "=== 00-setup.sh: bootstrapping project environment ==="
echo "Project root: $PROJ_ROOT"
echo ""

# ── Create data directory scaffold ───────────────────────────────────────────
echo "--- Creating data directories ---"
mkdir -p data/raw data/processed data/clean
echo "✓ data/raw"
echo "✓ data/processed"
echo "✓ data/clean"
echo ""

# ── Install Python environment via uv ────────────────────────────────────────
echo "--- Installing Python environment (uv sync) ---"
uv sync
echo "✓ Python environment ready"
echo ""

# ── Environment diagnostics ───────────────────────────────────────────────────
echo "--- Environment ---"
echo "bash:    $BASH_VERSION"
echo "uv:      $(uv --version)"
echo "python:  $(uv run python --version)"
echo "git:     $(git --version)"
echo ""
echo "=== 00-setup.sh complete ==="
```

Save (<kbd>Ctrl</kbd>+<kbd>x</kbd><kbd>Ctrl</kbd>+<kbd>s</kbd>) and quit
(<kbd>Ctrl</kbd>+<kbd>x</kbd><kbd>Ctrl</kbd>+<kbd>c</kbd>).

### 12b. Run and validate

```
bash scripts/00-setup.sh
```

```{bash verify-setup-sh, error=TRUE}
answer=$(bash scripts/00-setup.sh 2>&1 | tail -3)
echo "$answer"
check_bash "$answer" "HASH_SETUP_SH_TAIL3_TBD"
```

#### Reporter's Block/Resolution Log — Step 12

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 13: Write and Run `01-obtain.sh`

`01-obtain.sh` downloads all 14 decades of baby name HTML files from the
SSA website. It is **idempotent**: files that already exist are skipped.

### 13a. The specific shell glob

We use a **specific glob** to match only the decade-pattern filenames:

```
names[12][890][0-9]0s.html
```

This matches names1880s through names2010s — exactly 14 files. A vague
glob like `names*.html` would also match any accidentally present extra
files and produce wrong counts. Specificity is a safety net.

### 13b. Open and write the script

```
mg scripts/01-obtain.sh
```

```bash
#!/usr/bin/env bash
set -euo pipefail

# 01-obtain.sh — Download raw SSA baby name HTML files
# DSC 011 Sp26 — DSC-011-Sp26-Baby-Names
#
# Downloads one HTML file per decade (1880s–2010s) from SSA.
# Idempotent: skips files that already exist.
# Sleeps 1 second between new downloads (polite to SSA's server).
#
# Usage (from any directory):
#   bash scripts/01-obtain.sh

PROJ_ROOT="$(cd "$(dirname "${BASH_SOURCE[0]}")/.." && pwd)"
RAW_DIR="$PROJ_ROOT/data/raw"

echo "=== 01-obtain.sh: downloading raw HTML files ==="

EXISTING=$(find "$RAW_DIR" -name 'names[12][890][0-9]0s.html' \
               2>/dev/null | wc -l | tr -d ' ')

if [[ "$EXISTING" -eq 14 ]]; then
    echo "All 14 HTML files already present — skipping downloads."
    exit 0
fi

echo "Downloading ($EXISTING of 14 already present)..."
echo ""

for decade in {188..201}0; do
    html="$RAW_DIR/names${decade}s.html"
    url="https://www.ssa.gov/oact/babynames/decades/names${decade}s.html"

    if [[ -f "$html" ]]; then
        echo "  skip: names${decade}s.html"
    else
        echo "  downloading names${decade}s.html ..."
        sleep 1
        wget --quiet --output-document="$html" "$url"
        echo "  saved:  names${decade}s.html"
    fi
done

N=$(find "$RAW_DIR" -name 'names[12][890][0-9]0s.html' \
        2>/dev/null | wc -l | tr -d ' ')
echo ""
echo "✓ $N HTML files in data/raw/"
echo "=== 01-obtain.sh complete ==="
```

### 13c. Run and validate

```
bash scripts/01-obtain.sh
```

`names1990s.html` (downloaded in Step 11) will be skipped. The other 13
will be fetched — this takes about 13 seconds.

```{bash verify-html-count, error=TRUE}
answer=$(find data/raw -name 'names[12][890][0-9]0s.html' \
             2>/dev/null | wc -l | tr -d ' ')
echo "HTML files in data/raw/: $answer"
check_bash "$answer" "8527a891e224136950ff32ca212b45bc93f69fbb801c3b1ebedac52775f99e61"
```

Verify the specific glob matches exactly the right files — no extras:

```{bash verify-html-glob, error=TRUE}
echo "Files matched by specific glob:"
ls data/raw/names[12][890][0-9]0s.html 2>/dev/null \
    || echo "(no files matched yet)"
```

#### Reporter's Block/Resolution Log — Step 13

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 14: Begin `02-scrub.sh` — Step 1 (HTML → Plain Text)

`02-scrub.sh` is the data processing pipeline. Today we write only
**Step 1**: automating the html2text conversion you did by hand in Step 11.

### 14a. Open and write the script

```
mg scripts/02-scrub.sh
```

```bash
#!/usr/bin/env bash
set -euo pipefail

# 02-scrub.sh — Convert, parse, and clean raw baby name data
# DSC 011 Sp26 — DSC-011-Sp26-Baby-Names
#
# Step 1: Convert HTML files to plain text via Python (html2text)
# Steps 2+: added in future sessions
#
# Usage (from any directory):
#   bash scripts/02-scrub.sh
# Requirements:
#   - 00-setup.sh must have been run (.venv/ with html2text installed)
#   - 01-obtain.sh must have been run (14 HTML files in data/raw/)

PROJ_ROOT="$(cd "$(dirname "${BASH_SOURCE[0]}")/.." && pwd)"
RAW_DIR="$PROJ_ROOT/data/raw"
PROC_DIR="$PROJ_ROOT/data/processed"

# ── Guard: require 01-obtain.sh to have run first ────────────────────────────
N_RAW=$(find "$RAW_DIR" -name 'names[12][890][0-9]0s.html' \
            2>/dev/null | wc -l | tr -d ' ')
if [[ "$N_RAW" -lt 14 ]]; then
    echo "ERROR: Expected 14 HTML files in data/raw/ but found $N_RAW." >&2
    echo "       Run 'bash scripts/01-obtain.sh' first." >&2
    exit 1
fi

# ── Step 1: Convert each HTML file to plain text ──────────────────────────────
echo "=== 02-scrub.sh Step 1: converting HTML to plain text ==="

for decade in {188..201}0; do
    html="$RAW_DIR/names${decade}s.html"
    text="$PROC_DIR/names${decade}s.txt"

    if [[ -f "$text" ]]; then
        echo "  skip: names${decade}s.txt already exists"
    else
        echo "  converting names${decade}s.html → names${decade}s.txt"
        uv run --project "$PROJ_ROOT" python \
            -c 'import sys, html2text; \
                print(html2text.html2text(sys.stdin.read()))' \
            < "$html" > "$text"
    fi
done

N_TXT=$(find "$PROC_DIR" -name 'names[12][890][0-9]0s.txt' \
            2>/dev/null | wc -l | tr -d ' ')
echo ""
echo "✓ $N_TXT text files in data/processed/"
echo "=== Step 1 complete ==="
```

### 14b. Run the script

```
bash scripts/02-scrub.sh
```

`names1990s.txt` (converted in Step 11) will be skipped. The other 13
will be converted.

### 14c. Validate the text file count

```{bash verify-txt-count, error=TRUE}
answer=$(find data/processed -name 'names[12][890][0-9]0s.txt' \
             2>/dev/null | wc -l | tr -d ' ')
echo "Text files in data/processed/: $answer"
check_bash "$answer" "HASH_TXT_COUNT_TBD"
```

### 14d. Validate total data lines across all files

A data line starts with a rank number. With 200 ranks per decade and 14
decades we expect 2800 total data lines:

```{bash verify-data-lines, error=TRUE}
echo "Counting data lines across all 14 text files..."
answer=$(grep -l "^ *[0-9]" \
             data/processed/names[12][890][0-9]0s.txt \
         2>/dev/null \
         | xargs grep -c "^ *[0-9]" 2>/dev/null \
         | awk -F: '{s+=$2} END{print s}' \
         | tr -d ' ')
echo "Total data lines: $answer"
echo "(Expected: 2800 — 200 per decade × 14 decades)"
check_bash "$answer" "HASH_TOTAL_DATA_LINES_TBD"
```

#### Reporter's Block/Resolution Log — Step 14

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 15: Make the Second Commit

```
git add scripts/00-setup.sh scripts/01-obtain.sh scripts/02-scrub.sh \
        pyproject.toml uv.lock
git status
git commit -m "Add pipeline scripts: 00-setup, 01-obtain, 02-scrub Step 1"
```

Read `git status` before committing — `data/` and `.venv/` should not
appear (they are gitignored).

```{bash verify-second-commit, error=TRUE}
answer=$(git log --oneline 2>/dev/null | wc -l | tr -d ' ')
echo "Commit count: $answer"
check_bash "$answer" "d4735e3a265e16eee03f59718b9b5d03019c07d8b6c51f90da3a666eec13ab35"
```

#### Reporter's Block/Resolution Log — Step 15

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Step 16: Write the README, Package for Distribution, and Demonstrate Reproducibility

### 16a. Write the README

```
mg README.md
```

```markdown
# DSC-011-Sp26-Baby-Names

Reproducible analysis of U.S. Social Security Administration (SSA)
baby name data across 14 decades (1880s–2010s).

Built in DSC 011: Computing and Statistical Programming, UC Merced,
Spring 2026.

## Prerequisites

- bash 5.1+
- git
- uv (https://docs.astral.sh/uv/getting-started/installation/)
- wget

## Quick Start

    git clone <repo-url>
    cd DSC-011-Sp26-Baby-Names
    bash scripts/00-setup.sh     # create data dirs + install Python env
    bash scripts/01-obtain.sh    # download 14 HTML files from SSA (~14s)
    bash scripts/02-scrub.sh     # convert HTML to plain text

## Project Structure

    scripts/
        00-setup.sh     Bootstrap: data dirs + uv sync
        01-obtain.sh    Download raw HTML from SSA (idempotent)
        02-scrub.sh     Convert and clean data (idempotent)
    data/
        raw/            HTML files from SSA (gitignored — run 01-obtain.sh)
        processed/      Intermediate plain text files (gitignored)
        clean/          Final clean outputs (gitignored)

## Design Principles

All scripts are idempotent (safe to re-run). All scripts use the
PROJ_ROOT pattern so they can be called from any directory. Data
is never committed — only the scripts that reproduce it are tracked.
```

### 16b. Commit the README

```
git add README.md
git commit -m "Add README with quick-start instructions"
```

```{bash verify-readme, error=TRUE}
[[ -f README.md ]] \
    && echo "✓ README.md exists" \
    || echo "✗ README.md not found — write and save it first"

commits=$(git log --oneline 2>/dev/null | wc -l | tr -d ' ')
echo "Total commits: $commits"
```

### 16c. Reproducibility demonstration: clean and reconstitute

This is the final proof that the project is truly reproducible. We delete
all generated data and reconstruct it from the committed scripts alone.

**What we are proving:** any collaborator on any machine can clone this
repository, run three scripts, and get identical outputs.

**Step 1 — Delete all generated data:**

```
rm -rf data/raw/ data/processed/ data/clean/ .venv/
```

```{bash verify-clean, error=TRUE}
echo "After cleanup:"
[[ -d .venv ]] \
    && echo "✗ .venv still exists" \
    || echo "✓ .venv removed"
n=$(find data/ -type f 2>/dev/null | wc -l | tr -d ' ')
[[ "$n" -eq 0 ]] \
    && echo "✓ data/ is empty — all generated files removed" \
    || echo "✗ data/ still has $n files"
```

**Step 2 — Reconstitute using only the scripts:**

```
bash scripts/00-setup.sh
bash scripts/01-obtain.sh
bash scripts/02-scrub.sh
```

**Step 3 — Validate that outputs are identical:**

```{bash verify-reconstitution, error=TRUE}
echo "=== Reconstitution validation ==="
echo ""

answer_html=$(find data/raw -name 'names[12][890][0-9]0s.html' \
                  2>/dev/null | wc -l | tr -d ' ')
echo "HTML files in data/raw/: $answer_html"
check_bash "$answer_html" \
    "8527a891e224136950ff32ca212b45bc93f69fbb801c3b1ebedac52775f99e61"

echo ""

answer_txt=$(find data/processed -name 'names[12][890][0-9]0s.txt' \
                 2>/dev/null | wc -l | tr -d ' ')
echo "Text files in data/processed/: $answer_txt"
check_bash "$answer_txt" "HASH_TXT_COUNT_TBD"

echo ""
echo "Final project tree:"
tree --charset ascii data/ scripts/ 2>/dev/null
echo ""
answer_tree=$(tree --charset ascii data/ scripts/ 2>/dev/null)
check_bash "$answer_tree" "HASH_TREE_FINAL_TBD"
```

#### Reporter's Block/Resolution Log — Step 16

| Block | Resolution |
|:------|:-----------|
|       |            |

---

## Reporter's Mini-Retrospective Summary

**What helped the team succeed today:**

*(write here)*

**What could make the next session run better:**

*(write here)*

---

::: {.callout-important}
## ★ LAB 10 SUBMISSION POINT

**Render this notebook and submit the HTML to CatCourses.**

- **Due:** 5:30 pm Wednesday April 9 / 4:30 pm Friday April 11
- **Late deadline:** Friday April 11 at 11:59 pm
- **What to submit:** the rendered `.html` file (not the `.qmd`)

Before submitting, confirm:

- [ ] Your name is in the `author:` field
- [ ] Step 8 catch-up shows `✓` for all items from Demo 29
- [ ] Step 9 shows `✓ CORRECT` for commit count = 1
- [ ] Step 13c shows `✓ CORRECT` for 14 HTML files
- [ ] Step 14c shows `✓ CORRECT` for 14 text files
- [ ] Step 15 shows `✓ CORRECT` for commit count = 2
- [ ] Step 16c reconstitution shows `✓ CORRECT` for HTML and text counts

Partial credit is awarded for notebooks with some `✗ INCORRECT` outputs.
:::

---

```{r session-info, echo=FALSE}
sessionInfo()
```
